WNBA vs NBA Earnings In Comparable Players

### 2022-12-05 Manuscript/Manuscript.tex
Update Manuscript.tex

\documentclass[12pt]{article}

@@

\begin{document}

   \begin{center}
       \vspace*{1cm}
       \Huge
       \textbf{How to Earn a Salary}

       \Large
       \vspace{0.5cm}
        A Closer Look at WNBA and NBA Player Salaries, the Discrepancies, and What Statistics Might Influence Them

       \vspace{1.5cm}

       \textbf{Colin Sullivan}

       \vfill

       \vfill

       \vspace{0.8cm}

        \Large    
       Department of Statistics\\
       University of Connecticut\\
       Storrs, CT\\

   \end{center}
\newpage

@@

\par
Using another data set from Paine’s Github, a Python script called WNBA\_Cleanup.py was written and used to combine salary data from the two sources and 
filter out the desired candidate players from the data set. This paper will focus on comparing players from the 2018 NBA and WNBA seasons, so that filter 
was applied. The script also filtered out players who played less than 5 games and those that did not have salary data, so as to make the set as 
standardized as possible and remove outliers. The final data set after filtration and combination was called WNBA\_Combined.csv and has data (the 
aforementioned statistical categories) for 166 WNBA players from the 2018 season.
\par
A corresponding NBA dataset was found, this time belonging to Mustafa Baris Camli on the data-sharing site Kaggle \citep{nba}. This data set contains a 
list of all players from the NBA in the 2017-18 season, with similar advanced statistics to the WNBA dataset. The statistics found in the datasets can be 
seen in the following table:
\newline
\par
\begin{tabular}{*5l} 
\toprule
Statistics From the Datasets \\\midrule

Common Statistics & WNBA Exclusive & NBA Exclusive\\ 
\hline
 Player Name  &  Team Net Rating & 3 Point Attempt Rate  \\
 Age &  Composite Rating & Offensive Box +/-      \\ 
 Position &  Wins Generated & Defensive Box +/-\\
 Team &  & Box +/-     \\         
 Minutes Played &  & VORP       \\
 Player Efficiency Rating & &      \\
 True Shooting \% & & \\
 Free Throw Rate & &\\
 Offensive Rebounding \% & & \\
 Total Rebounding \% & & \\
 Assist \% & & \\
 Steal \% & & \\
 Block \% & & \\
 Turnover \% & & \\
 Usage \% & & \\
 Offensive Win Shares & & \\
 Defensive Win Shares  & & \\
 Win Shares & & \\
 Win Shares per 40 games & & \\
 Salary & & \\

 \hline
\end{tabular}
\newline
\par
The data set notably did not contain player salaries, so a second NBA data set was necessary. One was found on GitHub, created by Erik Gregory Webb, which 
contains the player salaries for every NBA player from 2000-2020 \citep{nba_salaries}. In a similar manner to the WNBA data, a Python script called 
NBA\_Cleanup.py was created to combine the data from the two datasets into one csv file. The data was then cleaned up by removing outliers (players 
playing under 5 games) and those that did not have salary data. The script was also used to combine data for players who played on multiple teams; the 
original data set would have players listed separately for each team they played on. The final data set after filtration and combination was called 
NBA\_Combined.csv and has data (the aforementioned statistical categories, plus player salaries) for 376 NBA players from the 2017-18 season.
\par
Finally, in order to compare the players accurately, the goal of checking their stats and salaries against league revenues must be achieved. As such, it 
was found that the NBA league revenue from 2018 was 8.01 billion USD \citep{NBA_Revenue} and the WNBA league revenue from the same year was 60 million USD 
\citep{WNBA_Revenue}.

\section{Research Design and Methods}
\hspace*{10mm}
Initially, we began by performing a paired sample t-test to confirm some expectations from the data. We do not need to perform a test to 
determine if NBA players are paid more than WNBA players, as that is commonly known and very obvious. As $\textit{Forbes}$ basketball author David Berri 
discovered, in general WNBA players are still paid significantly less than NBA players proportional to league revenue \citep{WNBA_Revenue}. In order to 
confirm that the data for the 2018 seasons fit with this mold, an initial t-test was performed:
$$H_0 = \mu_1 \leq \mu_0, H_a = \mu_1 > \mu_0$$
where $\mu_0$ represents the average WNBA salary proportional to league revenue and $\mu_a$ represents the same for the NBA. The test was done using a 
significance level of  5\%, or an alpha value of 0.05, and degrees of freedom $n_1 + n_2 - 2 = 516$. By the t-statistic chart, it must be checked if the 
found t-statistic is greater than 1.646. Using a Python script, Comp\_t\_tests.py, it was found that our t-statistic was 16.56. Since we want to convert 
this into a one tailed test, we halve the t-statistic to get 8.28, which is far greater than 1.646, so it can safely be said with 95\% confidence that our 
data follows the trend found in Berri's article.
\par
The primary focus of the analysis section was to analyze correlation using linear regression between advanced statistics and salary. This was done once 
again using python scripts: LR\_Test\_WNBA and LR\_Test\_NBA, for the WNBA and NBA, respectively, were used to read the data from the cleaned datasets and 
conduct simple linear regression analysis on all the advanced stats against player salaries. Then, using the scipy and matplotlib Python packages, each 
comparison was graphed with a line of best fit and a correlation coefficient ($R^2$) was found. This correlation coefficient tells us how strong the 
correlation effect is between the statistic and salary, which will help us draw significantly more accurate conclusions from the data then if simply the 
slope was examined. There were twelve overlapping statistics between the WNBA and NBA datasets, and the correlation coefficients for these statistics in 
both leagues are shown below, with the standard error in parenthesis:

\begin{tabular}{*5l}    \toprule
WNBA vs NBA Correlation & Coefficients For & Selected Advanced Statistics \\\midrule

Statistic 
&  CC: WNBA   & CC: NBA \\

 PER  &  0.019458 (1.576)  & 0.25444 (0.014) \\
 FTr   &  0.001641 (0.036)  & 0.033096 (0.009)       \\ 
 ORB\%   &  0.019717 (0.929)  & 0.000274 (0.273)       \\   
 TRB\%   &  0.000017 (1.080)   & 0.007801 (0.337)       \\
 AST\%   &  0.041707 (1.936)  & 0.110984 (0.608)    \\
 STL\%  &  0.003666 (0.175)  & 0.006126 (0.045)       \\    
 BLK\%  & 0.000444 (0.337)   & 0.000141 (0.114)     \\
 TOV\%  & 0.010470 (1.432)  & 0.004207 (0.315)    \\
 USG\%   & 0.033326 (1.146)   & 0.202527 (0.350)       \\
 OWS   & 0.117693 (0.276)   & 0.337278 (0.121)        \\
 DWS  & 0.151183 (0.133)   & 0.234252 (0.068)         \\
 WS   &  0.153999 (0.365)  & 0.364165 (0.164)                 \\
 (WS/40)/(WS/48)   &  0.047009 (0.022)  & 0.151640 (0.004) \\
 \hline
 Sample Size & 142 & 376
\end{tabular}
\newline
\par
The graphs for the strongest correlation in both leagues (both the win shares statistic) are shown below:
\newline
\includegraphics[width=.85\textwidth]{NBA_WS_Graph.png}
\newline
\includegraphics[width=.85\textwidth]{WNBA_WS_Graph.png}

\par
Finally, within the Comp\_t\_tests.py Python script, analysis was conducted on the salaries of the top players in each league. Instead of looking at 
simple statistics, work that has been done by DiCicio \citep{Hoop_Dreams}, this method used the advanced statistics from the data set. Six statistics were selected: Minutes 
Played (MP), Player Efficiency Rating (PER), Free Throw rate (FTr), Offensive Win Shares (OWS), Defensive Win Shares (DWS), and Win Shares (WS). The 
script was written to sort the datasets by these statistics, and the top 10 players in each league in these statistics were extracted. Then the salaries 
were compared to each respective league's median salary, and the average of these top 10 players relations to the average were compiled. The entries in 
the graph can be interpreted as follows: If the value is greater than 1, than the top 10 players in that statistic are paid the value as a factor above 
league average. If the value is less than 1, than the top 10 players in that statistic are paid the value as a factor below league average. The chart is 
below:
\vspace*{1cm}
\begin{tabular}{*5l} 
\toprule
How Much Compared to Avg & the Top 10 Players Are Paid & For Advanced Stats \\\midrule

Statistic &  WNBA  & NBA \\

 MP  &  1.445  & 5.353 \\
 PER   &  1.254   & 5.608        \\       
 FTr   &  1.087   & 2.053       \\         
 OWS   & 1.445    & 5.666        \\
 DWS  & 1.435    & 3.462       \\
 WS   &  1.530  & 5.446                 \\
 \hline
\end{tabular}

@@

shown in the aforementioned t-tests), it seemed most appropriate to simply see which statistics correlated the strongest to salary increases. Table 1 

@@

The visual distribution of pay vs performance in the graphs is particularly interesting. While the correlation coefficient for both the WNBA and the NBA 

@@

the trend line. The WNBA graph, however, looks much less traditional. It forms a sort of "U" shape, where players are almost all clustered into two salary 

@@

Finally, some data manipulation was done to find a more simple statistic to try and determine how excellence in a statistic impacts pay. As Table 2 shows, 

@@

\section{Conclusion}
\hspace*{10mm}
The motivation of this paper was to try and determine what goes into the pay gap in men’s and women’s basketball and primarily to see if the salary 
models were similar (just dramatically scaled down in the much poorer league). It was confirmed that the data for the 2018 season fit with precedent 
in terms of NBA salaries being much higher, both base and proportional to league revenue. The regression analysis revealed that the correlation 
coefficients for NBA salaries to selected impactful advanced statistics was in general much higher than that of the WNBA, revealing that the 
aforementioned salary models were dramatically different. It was, however, determined that finishing in the top 10 in said statistics resulted in pay 
increases over league median, however the NBA’s pay increases were significantly more due to their much larger spread in salaries. 
\par
Perhaps one of the most interesting conclusions is drawn when examining the graphs of NBA and WNBA pay distribution against Win Shares performance. 
Not only is there much less pay increase opportunity in the WNBA, but it seems as though almost all players are grouped into one of two pay rates, 
which may or may not be based solely upon performance.
\par
Next steps from this data involve researching what goes into league revenue and attempting to determine why it has never been profitable. 
It is important to acknowledge one last time that the WNBA is hamstrung by this lack of revenue, but the contract structures and much lower 
pay proportional to league revenue reinforce male dominant stereotypes in an already masculine world with a widespread gender pay gap in nearly
every industry. Ultimately, the NBA players get rewarded much more for performance than the WNBA, which may motivate them to improve more and bring 
more excitepment and popularity to the league. With less monetary motivation to perform, the WNBA continues to sell itself well short of its potential 
and does not give the players the respect that they deserve.

>>>>

\documentclass[12pt, titlepage]{article}

@@

\begin{document}
\maketitle

%   \begin{center}
%       \vspace*{1cm}
%       \Huge
%       \textbf{How to Earn a Salary}
%
%       \Large
%       \vspace{0.5cm}
%        A Closer Look at WNBA and NBA Player Salaries, the Discrepancies, and What Statistics Might Influence Them
%            
%       \vspace{1.5cm}
%
%       \textbf{Colin Sullivan}
%       
%       \vfill
%
%
%       \vfill
%            
%            
%       \vspace{0.8cm}
%     
%         
%        \Large    
%       Department of Statistics\\
%       University of Connecticut\\
%       Storrs, CT\\
%            
%   \end{center}
\newpage

@@

\par
Using another data set from Paine’s Github, a Python script called WNBA\_Cleanup.py was written and used to combine salary data from the two sources and 
filter out the desired candidate players from the data set. This paper will focus on comparing players from the 2018 NBA and WNBA seasons, so that filter 
was applied. The script also filtered out players who played less than 5 games and those that did not have salary data, so as to make the set as 
standardized as possible and remove outliers. The final data set after filtration and combination was called WNBA\_Combined.csv and has data (the 
aforementioned statistical categories) for 166 WNBA players from the 2018 season.
\par
A corresponding NBA dataset was found, this time belonging to Mustafa Baris Camli on the data-sharing site Kaggle \citep{nba}. This data set contains a 
list of all players from the NBA in the 2017-18 season, with similar advanced statistics to the WNBA dataset. The statistics found in the datasets can be 
seen in \ref{tab:stats}.

\begin{table}[tbp]
    \caption{Statistics From the Datasets}
    \label{tab:stats}
\centering
\begin{tabular}{crr}
\toprule

Common Statistics & WNBA Exclusive & NBA Exclusive\\ 
\midrule
 Player Name  &  Team Net Rating & 3 Point Attempt Rate  \\
 Age &  Composite Rating & Offensive Box +/-      \\ 
 Position &  Wins Generated & Defensive Box +/-\\
 Team &  & Box +/-     \\         
 Minutes Played &  & VORP       \\
 Player Efficiency Rating & &      \\
 True Shooting \% & & \\
 Free Throw Rate & &\\
 Offensive Rebounding \% & & \\
 Total Rebounding \% & & \\
 Assist \% & & \\
 Steal \% & & \\
 Block \% & & \\
 Turnover \% & & \\
 Usage \% & & \\
 Offensive Win Shares & & \\
 Defensive Win Shares  & & \\
 Win Shares & & \\
 Win Shares per 40 games & & \\
 Salary & & \\

 \bottomrule
\end{tabular}
\end{table}

\par
The data set notably did not contain player salaries, so a second NBA data set was necessary. One was found on GitHub, created by Erik Gregory Webb, which 
contains the player salaries for every NBA player from 2000-2020 \citep{nba_salaries}. In a similar manner to the WNBA data, a Python script called 
NBA\_Cleanup.py was created to combine the data from the two datasets into one csv file. The data was then cleaned up by removing outliers (players 
playing under 5 games) and those that did not have salary data. The script was also used to combine data for players who played on multiple teams; the 
original data set would have players listed separately for each team they played on. The final data set after filtration and combination was called 
NBA\_Combined.csv and has data (the aforementioned statistical categories, plus player salaries) for 376 NBA players from the 2017-18 season.
\par
Finally, in order to compare the players accurately, the goal of checking their stats and salaries against league revenues must be achieved. As such, it 
was found that the NBA league revenue from 2018 was 8.01 billion USD \citep{NBA_Revenue} and the WNBA league revenue from the same year was 60 million USD 
\citep{WNBA_Revenue}.

\section{Research Design and Methods}
\hspace*{10mm}
Initially, we began by performing a paired sample t-test to confirm some expectations from the data. We do not need to perform a test to 
determine if NBA players are paid more than WNBA players, as that is commonly known and very obvious. As $\textit{Forbes}$ basketball author David Berri 
discovered, in general WNBA players are still paid significantly less than NBA players proportional to league revenue \citep{WNBA_Revenue}. In order to 
confirm that the data for the 2018 seasons fit with this mold, an initial t-test was performed:
$$H_0 = \mu_1 \leq \mu_0, H_a = \mu_1 > \mu_0$$
where $\mu_0$ represents the average WNBA salary proportional to league revenue and $\mu_a$ represents the same for the NBA. The test was done using a 
significance level of  5\%, or an alpha value of 0.05, and degrees of freedom $n_1 + n_2 - 2 = 516$. By the t-statistic chart, it must be checked if the 
found t-statistic is greater than 1.646. Using a Python script, Comp\_t\_tests.py, it was found that our t-statistic was 16.56. Since we want to convert 
this into a one tailed test, we halve the t-statistic to get 8.28, which is far greater than 1.646, so it can safely be said with 95\% confidence that our 
data follows the trend found in Berri's article.
\par
The primary focus of the analysis section was to analyze correlation using linear regression between advanced statistics and salary. This was done once 
again using python scripts: LR\_Test\_WNBA and LR\_Test\_NBA, for the WNBA and NBA, respectively, were used to read the data from the cleaned datasets and 
conduct simple linear regression analysis on all the advanced stats against player salaries. Then, using the scipy and matplotlib Python packages, each 
comparison was graphed with a line of best fit and a correlation coefficient ($R^2$) was found. This correlation coefficient tells us how strong the 
correlation effect is between the statistic and salary, which will help us draw significantly more accurate conclusions from the data then if simply the 
slope was examined. There were twelve overlapping statistics between the WNBA and NBA datasets, and the correlation coefficients for these statistics in 
both leagues are shown in \ref{tab:main_data}, with the standard error in parenthesis.

\begin{table}[tbp]
    \caption{WNBA vs NBA Correlation Coefficients For Selected Advanced Statistics}
    \label{tab:main_data}
\centering
\begin{tabular}{crr}
\toprule

Statistic &  CC: WNBA   & CC: NBA \\ 
\midrule
 PER  &  0.019458 (1.576)  & 0.25444 (0.014) \\
 FTr   &  0.001641 (0.036)  & 0.033096 (0.009)       \\ 
 ORB\%   &  0.019717 (0.929)  & 0.000274 (0.273)       \\   
 TRB\%   &  0.000017 (1.080)   & 0.007801 (0.337)       \\
 AST\%   &  0.041707 (1.936)  & 0.110984 (0.608)    \\
 STL\%  &  0.003666 (0.175)  & 0.006126 (0.045)       \\    
 BLK\%  & 0.000444 (0.337)   & 0.000141 (0.114)     \\
 TOV\%  & 0.010470 (1.432)  & 0.004207 (0.315)    \\
 USG\%   & 0.033326 (1.146)   & 0.202527 (0.350)       \\
 OWS   & 0.117693 (0.276)   & 0.337278 (0.121)        \\
 DWS  & 0.151183 (0.133)   & 0.234252 (0.068)         \\
 WS   &  0.153999 (0.365)  & 0.364165 (0.164)                 \\
 (WS/40)/(WS/48)   &  0.047009 (0.022)  & 0.151640 (0.004) \\
 \hline
 Sample Size & 142 & 376\\
 \bottomrule
\end{tabular}
\end{table}

\par
The graphs for the strongest correlation in both leagues (both the win shares statistic) are shown in figures \ref{fig:NBA_WS} and \ref{fig:WNBA_WS}.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=.75\textwidth]{NBA_WS_Graph.png}
  \caption{NBA Player Salaries vs Win Share Performance.}
  \label{fig:NBA_WS}
\end{figure}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.75\textwidth]{WNBA_WS_Graph.png}
  \caption{WNBA Player Salaries vs Win Share Performance.}
  \label{fig:WNBA_WS}
\end{figure}

\par
Finally, within the Comp\_t\_tests.py Python script, analysis was conducted on the salaries of the top players in each league. Instead of looking at 
simple statistics, work that has been done by DiCicio \citep{Hoop_Dreams}, this method used the advanced statistics from the data set. Six statistics were selected: Minutes 
Played (MP), Player Efficiency Rating (PER), Free Throw rate (FTr), Offensive Win Shares (OWS), Defensive Win Shares (DWS), and Win Shares (WS). The 
script was written to sort the datasets by these statistics, and the top 10 players in each league in these statistics were extracted. Then the salaries 
were compared to each respective league's median salary, and the average of these top 10 players relations to the average were compiled. The entries in 
the graph can be interpreted as follows: If the value is greater than 1, than the top 10 players in that statistic are paid the value as a factor above 
league average. If the value is less than 1, than the top 10 players in that statistic are paid the value as a factor below league average. The chart is 
shown in \ref{tab:top10}.

\begin{table}[tbp]
    \caption{How Much Compared to Avg the Top 10 Players Are Paid For Advanced Stats.}
    \label{tab:top10}
\centering
\begin{tabular}{crr}
\toprule
Statistic &  WNBA  & NBA \\ 
\midrule
 MP  &  1.445  & 5.353 \\
 PER   &  1.254   & 5.608        \\       
 FTr   &  1.087   & 2.053       \\         
 OWS   & 1.445    & 5.666        \\
 DWS  & 1.435    & 3.462       \\
 WS   &  1.530  & 5.446        \\
 \bottomrule
\end{tabular}
\end{table}

@@

shown in the aforementioned t-tests), it seemed most appropriate to simply see which statistics correlated the strongest to salary increases. \ref{tab:main_data} 

@@

The visual distribution of pay vs performance in figures \ref{fig:NBA_WS} and \ref{fig:WNBA_WS} is particularly interesting. While the correlation coefficient for both the WNBA and the NBA 

@@

the trend line. The WNBA figure \ref{fig:WNBA_WS}, however, looks much less traditional. It forms a sort of "U" shape, where players are almost all clustered into two salary 

@@

Finally, some data manipulation was done to find a more simple statistic to try and determine how excellence in a statistic impacts pay. As \ref{tab:top10} shows, 

@@

\section{Conclusion}
\hspace*{10mm}
The motivation of this paper was to try and determine what goes into the pay gap in men’s and women’s basketball and primarily to see if the salary 
models were similar (just dramatically scaled down in the much poorer league). It was confirmed that the data for the 2018 season fit with precedent 
in terms of NBA salaries being much higher, both base and proportional to league revenue. The regression analysis revealed that the correlation 
coefficients for NBA salaries to selected impactful advanced statistics was in general much higher than that of the WNBA, revealing that the 
aforementioned salary models were dramatically different. It was, however, determined that finishing in the top 10 in said statistics resulted in pay 
increases over league median, however the NBA’s pay increases were significantly more due to their much larger spread in salaries. 
\par
Perhaps one of the most interesting conclusions is drawn when examining figures \ref{fig:NBA_WS} and \ref{fig:WNBA_WS} of NBA and WNBA pay distribution against Win Shares performance. 
Not only is there much less pay increase opportunity in the WNBA, but it seems as though almost all players are grouped into one of two pay rates, 
which may or may not be based solely upon performance.
\par
Next steps from this data involve researching what goes into league revenue and attempting to determine why it has never been profitable. 
It is important to acknowledge one last time that the WNBA is hamstrung by this lack of revenue, but the contract structures and much lower 
pay proportional to league revenue reinforce male dominant stereotypes in an already masculine world with a widespread gender pay gap in nearly
every industry. Ultimately, the NBA players get rewarded much more for performance than the WNBA, which may motivate them to improve more and bring 
more excitepment and popularity to the league. With less monetary motivation to perform, the WNBA continues to sell itself well short of its potential 
and does not give the players the respect that they deserve.
